Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-03-28 submission/manuscript.Rmd
Incorporate figure legends

regression <- read_tsv("../data/rf/regression_data_pool.tsv",
												col_types=cols(class=col_character(),
													input=col_character(),
													.default=col_double()
												)
 											) %>%
							mutate(microbiome=str_replace(input, "fit_", ""),
										microbiome=str_replace(microbiome, "_fit", "")
								) %>%
							select(class, microbiome, test_Rsquared) %>%
							filter(class %in% scfa_types) %>%
							filter(microbiome %in% regression_features) %>%
							mutate(data_type = case_when(
					       microbiome %in% c("genus", "otu")  ~ "16S",
								 microbiome %in% c("kegg", "opf")  ~ "metagenomics",
								 microbiome %in% c("pc2ec", "pc2pathways")  ~ "picrust"
							 )) %>%
							group_by(class, microbiome, data_type) %>%
							summarize(median_Rsquared = median(test_Rsquared))

@@

**Random Forest models.**

@@

\includegraphics{../results/figures/scfa_comparisons.pdf}

**Figure 1. SCFA concentrations did not vary meaningfully with diagnosis of colonic lesions or with treatment for adenomas or carcinomas.** (A) We measured the concentration of fecal SCFAs from individuals with normal colons (N=`r n_normal`) or those with adenoma (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) was for isobutyrate. (B) A subset of individuals diagnosed with adenomas (N=`r n_follow_adenoma`) or carcinomas (N=`r n_follow_cancer`) who underwent treatment were resampled a year after the initial sampling; one extreme propionate value (124.4 mmol/kg) was included in the adenoma analysis but censored from the visualization for clarity.

@@

\includegraphics{../results/figures/scfa_modelling.pdf}

**Figure 2. SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions. 16S rRNA gene and metagenomic sequence data do not predict SCFAs concentrations.**

@@

\includegraphics{../results/figures/classification_testing.pdf}

**Figure S1. Comparison of training and testing results.**

@@

\includegraphics{../results/figures/regression_testing.pdf}

**Figure S1. Comparison of training and testing results.**

>>>>

regression <- read_tsv("../data/rf/regression_data_pool.tsv",
												col_types=cols(class=col_character(),
													input=col_character(),
													.default=col_double()
												)
 											) %>%
							mutate(microbiome=str_replace(input, "fit_", ""),
										microbiome=str_replace(microbiome, "_fit", "")
								) %>%
							select(class, microbiome, test_Rsquared) %>%
							filter(class %in% scfa_types) %>%
							filter(microbiome %in% regression_features) %>%
							mutate(data_type = case_when(
					       microbiome %in% c("genus", "otu")  ~ "16S",
								 microbiome %in% c("kegg", "opf")  ~ "metagenomics",
								 microbiome %in% c("pc2ec", "pc2pathways")  ~ "picrust"
							 )) %>%
							group_by(class, microbiome, data_type) %>%
							summarize(median_Rsquared = median(test_Rsquared)) %>%
							ungroup()

@@

**Random Forest models.** The classification models were built to predict lesion type from microbiome information with or without SCFA concentrations. The regression models were built to predict the SCFA concentrations of acetate, butyrate, and propionate from microbiome information. For classification and regression models, we pre-processed the features by scaling them to vary between zero and one. Features with no variance in the training set were removed from both the training and testing sets. We randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining data. The training set was used for hyperparameter selection and training the model, and the test set was used for evaluating prediction performance. For each model, the best performing hyperparameter, mtry, was selected in an internal five-fold cross-validation of the training set with 100 randomizations. Six values of mtry were tested and the value that provided the largest AUROC or R^2^ was selected. We trained the random forest model using the selected mtry value and predicted the held-out test set. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance. We used AUROC and R^2^ as the prediction performance metric for classification and regression models, respectively. We used randomForest package implemented to the caret package (version `r installed.packages()["randomForest" ,"Version"]`) in R statistical software (version `r installed.packages()["caret" ,"Version"]`) for our models.

@@

\includegraphics{figure_1.png}

**Figure 1. SCFA concentrations did not vary meaningfully with diagnosis of colonic lesions or with treatment for adenomas or carcinomas.** (A) We measured the concentration of fecal SCFAs from individuals with normal colons (N=`r n_normal`) or those with adenoma (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`). (B) A subset of individuals diagnosed with adenomas (N=`r n_follow_adenoma`) or carcinomas (N=`r n_follow_cancer`) who underwent treatment were resampled a year after the initial sampling; one extreme propionate value (124.4 mmol/kg) was included in the adenoma analysis but censored from the visualization for clarity.

@@

\includegraphics{figure_2.png}

**Figure 2. SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions and cannot be reliably predicted from 16S rRNA gene or metagenomic sequence data.** (A) The median AUROC for diagnosing individuals as having adenomas or carcinomas using SCFAs was slightly better than than chance (depicted by horizontal line at 0.50), but did improve performance of the models generated using 16S rRNA gene sequence data. (B) Regression models that were trained using 16S rRNA gene sequence, metagenomic, and PICRUSt data to predict the concentrations of SCFAs performed poorly (all median R^2^ values < `r max(regression$median_Rsquared)`). Regression models generated using 16S rRNA gene sequence and PICRUSt data included data from `r sum(n$n)` samples and those generated using metagenomic data included data from `r sum(n_s$n)` samples.

@@

\includegraphics{figure_s1.png}

**Figure S1. Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** Random Forest classification models were generated to differentiate between subjects with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the four SCFAs as additional features. Random Forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples. The training data was used to fit the mtry hyperparameter using 100 five-fold cross validations for six values of mtry. The mtry with the largest AUROC was selected to train the model, which was then tested on the held out 20% of the samples.

@@

\includegraphics{figure_s2.png}

**Figure S2. Comparison of training and testing results for regression models shows that the models are robust and are not overfit.** Random Forest regression models were generated to predict the concentration of each SCFA using each subjects' microbiome data generated using 16S rRNA gene sequence and metagenomic sequence data. These regression models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples. The training data was used to fit the mtry hyperparameter using 100 five-fold cross validations for six values of mtry. The mtry with the largest R^2^ was selected to train the model, which was then tested on the held out 20% of the samples.

### 2019-03-28 submission/manuscript.tex
Incorporate figure legends

\textbf{Random Forest models.}

@@

\includegraphics{../results/figures/scfa_comparisons.pdf}

\textbf{Figure 1. SCFA concentrations did not vary meaningfully with
diagnosis of colonic lesions or with treatment for adenomas or
carcinomas.} (A) We measured the concentration of fecal SCFAs from
individuals with normal colons (N=172) or those with adenoma (N=198) or
carcinomas (N=120) was for isobutyrate. (B) A subset of individuals
diagnosed with adenomas (N=41) or carcinomas (N=26) who underwent
treatment were resampled a year after the initial sampling; one extreme
propionate value (124.4 mmol/kg) was included in the adenoma analysis
but censored from the visualization for clarity.

@@

\includegraphics{../results/figures/scfa_modelling.pdf}

\textbf{Figure 2. SCFA concentrations do not improve models for
diagnosing the presence of adenomas, carcinomas, or all lesions. 16S
rRNA gene and metagenomic sequence data do not predict SCFAs
concentrations.}

@@

\includegraphics{../results/figures/classification_testing.pdf}

\textbf{Figure S1. Comparison of training and testing results.}

@@

\includegraphics{../results/figures/regression_testing.pdf}

\textbf{Figure S1. Comparison of training and testing results.}

>>>>

\textbf{Random Forest models.} The classification models were built to
predict lesion type from microbiome information with or without SCFA
concentrations. The regression models were built to predict the SCFA
concentrations of acetate, butyrate, and propionate from microbiome
information. For classification and regression models, we pre-processed
the features by scaling them to vary between zero and one. Features with
no variance in the training set were removed from both the training and
testing sets. We randomly split the data into training and test sets so
that the training set consisted of 80\% of the full dataset while the
test set was composed of the remaining data. The training set was used
for hyperparameter selection and training the model, and the test set
was used for evaluating prediction performance. For each model, the best
performing hyperparameter, mtry, was selected in an internal five-fold
cross-validation of the training set with 100 randomizations. Six values
of mtry were tested and the value that provided the largest AUROC or
R\textsuperscript{2} was selected. We trained the random forest model
using the selected mtry value and predicted the held-out test set. The
data-split, hyperparameter selection, training and testing steps were
repeated 100 times to get a reliable and robust reading of model
prediction performance. We used AUROC and R\textsuperscript{2} as the
prediction performance metric for classification and regression models,
respectively. We used randomForest package implemented to the caret
package (version 4.6-14) in R statistical software (version 6.0-81) for
our models.

@@

\includegraphics{figure_1.png}

\textbf{Figure 1. SCFA concentrations did not vary meaningfully with
diagnosis of colonic lesions or with treatment for adenomas or
carcinomas.} (A) We measured the concentration of fecal SCFAs from
individuals with normal colons (N=172) or those with adenoma (N=198) or
carcinomas (N=120). (B) A subset of individuals diagnosed with adenomas
(N=41) or carcinomas (N=26) who underwent treatment were resampled a
year after the initial sampling; one extreme propionate value (124.4
mmol/kg) was included in the adenoma analysis but censored from the
visualization for clarity.

@@

\includegraphics{figure_2.png}

\textbf{Figure 2. SCFA concentrations do not improve models for
diagnosing the presence of adenomas, carcinomas, or all lesions and
cannot be reliably predicted from 16S rRNA gene or metagenomic sequence
data.} (A) The median AUROC for diagnosing individuals as having
adenomas or carcinomas using SCFAs was slightly better than than chance
(depicted by horizontal line at 0.50), but did improve performance of
the models generated using 16S rRNA gene sequence data. (B) Regression
models that were trained using 16S rRNA gene sequence, metagenomic, and
PICRUSt data to predict the concentrations of SCFAs performed poorly
(all median R\textsuperscript{2} values \textless{} 0.14). Regression
models generated using 16S rRNA gene sequence and PICRUSt data included
data from 490 samples and those generated using metagenomic data
included data from 78 samples.

@@

\includegraphics{figure_s1.png}

\textbf{Figure S1. Comparison of training and testing results for
classification models shows that the models are robust and are not
overfit.} Random Forest classification models were generated to
differentiate between subjects with normal colons and those with
adenomas or carcinomas using 16S rRNA gene sequence data that were
clustered into genera or OTUs with and without including the four SCFAs
as additional features. Random Forest classification models were
generated by partitioning the samples into a training set with 80\% of
the data and a testing set with the remaining samples. The training data
was used to fit the mtry hyperparameter using 100 five-fold cross
validations for six values of mtry. The mtry with the largest AUROC was
selected to train the model, which was then tested on the held out 20\%
of the samples.

@@

\includegraphics{figure_s2.png}

\textbf{Figure S2. Comparison of training and testing results for
regression models shows that the models are robust and are not overfit.}
Random Forest regression models were generated to predict the
concentration of each SCFA using each subjects' microbiome data
generated using 16S rRNA gene sequence and metagenomic sequence data.
These regression models were generated by partitioning the samples into
a training set with 80\% of the data and a testing set with the
remaining samples. The training data was used to fit the mtry
hyperparameter using 100 five-fold cross validations for six values of
mtry. The mtry with the largest R\textsuperscript{2} was selected to
train the model, which was then tested on the held out 20\% of the
samples.
